\large How resource abundance and stochasticity affect organisms' spatial needs

### 2023-09-08 writing/manuscript.Rmd
updated based on Chris's comments

---
title: \large How resource abundance and stochasticity affect animals' spatial needs
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Simon Wood
    institute: uofe
  # - name: Adam T. Ford
  #   institute: biol
  #   institute: braes
  # - name: Jacob R. Goheen
  #   institute: wyom
  - name: E. Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - uofe: School of Mathematics, James Clerk Maxwell Building, University of Edinburgh, Edinburgh, Unted Kingdom.
  # - wyom: Department of Zoology and Physiology, University of Wyoming, Laramie, Wyoming, United States of America.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
    -  \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

<!-- define custom math symbols -->

\newcommand*\e{\mathbb E} <!-- expectation (blackboard bold) E -->

\newcommand*\var{\mathbb V} <!-- variance (blackboard bold) V -->

@@

#' place figures before the paragraph they are referred to so `wrapfigure` places them
#' beside the paragraph
```

@@

\noindent The amount of space animals use is thought to be tightly linked to the availability of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the *mean* amount of resources, while responses to the variance around the mean are still largely unknown. This is not a trivial oversight. Animals adjust to variable environmental conditions, so failing to account for the effects of resource unpredictability can result in a limited understanding of animals' spatial needs, challenging ecological theory and applied conservation alike. In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance. We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on animals' space use. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in animal space use. Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado <!--an impala (*Aepyceros melampus*), and multiple moose (*Alces alces*)--> to show how this framework can be applied to better understand the movement ecology of free-ranging animals. Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.

@@

The relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested that change in resource abundance is a strong determinant of how animals decide where to live and when to reproduce. Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an animal's home-range size is

@@

\noindent where $H$ is the animal's **home-range size**, $C$ is the animal's resource consumption, and $R$ is the **resources** the animal can access. Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but a species' spatial requirements likely depend on numerous other factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020]. While many researchers have since demonstrated that animals adapt their home ranges in response to resources abundance, few studies build upon others' previous work. Instead, results are reported as independent, novel findings. Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on animals' spatial needs. In addition, while much work has been done on estimating animals' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.

In this paper, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an animal might use. For instance, a grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources. In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement and life history strategies being selected for. Although it is generally expected that resource unpredictability will decrease animals' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

Here, we illustrate how animals' spatial needs depend on both the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation. Next, we provide a review of the effects of resource *abundance* on animals' spatial needs while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource *stochasticity* on animals' spatial needs while suggesting a second unifying hypothesis. Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022]. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

# Concepts and definitions

<!--\noindent In this section, we presents the foundational concepts and terms used in this manuscript. We introduce each concept without assuming a statistical or quantitative background to maximize accessibility and minimize misunderstandings and misinterpretations.-->

## Resources as a random variable

\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Resource abundance is often unpredictable (and difficult to quantify), since it depends on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, we can let the random variable $R$ indicate the resource abundance at a given point in space and time. Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.

### Probability distributions

\noindent Random variables are defined by specifying the **distribution** the variable is assumed to follow (e.g., Gaussian, Gamma, Poisson, Bernoulli). Since the variable is random, it can take multiple possible values, each with different probabilities. The set or range of values which have non-zero probabilities in a distribution is referred to as the distribution's **support**.

There are many distributions we can assign to $R$ depending on how we quantify it. For instance, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included). However, there is no commonly used distribution with that support, so the best option is to rescale NDVI to (0, 1) and use a beta distribution (see the section below on applying this framework). Alternatively, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support positive real numbers, such as a Gamma or log-normal distribution. If $R$ is a discrete integer variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.

### Expected resource abundance, $\e(R)$

\noindent Since the exact value of $R$ at a given time and location is unknown, comparing the magnitude in $R$ between two locations or time periods requires a quantitative measure of what value we believe $R$ will take, on average. The **expectation** of a random variable is the value one can *expect* the random variable to take, on average, in the long term. Thus, we can let $\e(R)$ indicate the expectation of the random variable $R$. When the mean changes over time, such as in strongly seasonal regions, we explicitly indicate that $\e(R)$ changes over time by writing $\e(R)$ as a function of time, $t$: $\e(R|t) = \mu(t)$. We indicate the *estimated* average resource abundance over time by adding a caret on the symbol: $\hat \mu(t)$.

### Variance in resource abundance, $\var(R)$

\noindent In viewing resources as a random variable, we not only obtain a formal framework for describing $\e(R)$, but also the spread around $\e(R)$. A random variable's **variance** is a measure of its unpredictability (i.e., variability). We use the notation $\var(R)$ to indicate the variance in $R$ after accounting for changes in $\mu(t)$, and we use $\sigma^2(t)$ to indicate its function over time (with estimate $\widehat{\sigma^2}(t)$). Consequently, while $R$ may vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others. Additionally, we distinguish between two different sources of variation. We refer to known or expected spatiotemporal differences or changes as **heterogeneity**, which we consider as a change in $\mu(t)$. In contrast, we use the term **stochasticity** for unpredictable and unexpected variation (i.e., due to $\sigma^2(t)$. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but it becomes predictable once a bear visits all patches in the valley. In contrast, whether next year will be a good or bad year for berries is stochastic because it depends on factors the bear cannot predict.

### Simulating resource abundance

\noindent In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa). As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically; in the example from above, a mean of zero kg of berries within the bear's habitat during winter implies the variance in berries is also zero.

# Effects of resource abundance, $\e(R)$

```{r mu-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of resource abundance ($\\e(R)$). We expect low values of $\\e(R)$ to result in high spatial needs as animals are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic. As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive. Note that the relationship between $\\e(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
```

```{r, eval=FALSE, echo=FALSE}
# to help visualize the consequences of different E(R) and V(R) if R ~ Gamma
dgamma2 <- function(x, mu, sigma2, lower.tail = FALSE) {
    # mean = k * theta
    # sigma^2 = k * theta^2
    dgamma(x = x,
           shape = mu^2 / sigma2, # (k * theta)^2 / (k * theta^2)
           scale = sigma2 / mu)   # (k * theta^2) / (k * theta)
}
x <- seq(0, 10, length.out = 250)
layout(matrix(1:4, ncol = 2, byrow = TRUE))
plot(x, dgamma2(x, mu = 1, sigma2 = 0.5), type = 'l', ylim = c(0, 0.8),
     xlab = NA, ylab = 'Density', main = 'low \U03BC, low \U03C3\U00B2')
plot(x, dgamma2(x, mu = 5, sigma2 = 0.5), type = 'l', ylim = c(0, 0.8),
     xlab = NA, ylab = 'Density', main = 'high \U03BC, low \U03C3\U00B2')
plot(x, dgamma2(x, mu = 1, sigma2 = 5), type = 'l', ylim = c(0, 0.8),
     xlab = NA, ylab = 'Density', main = 'low \U03BC, high \U03C3\U00B2')
plot(x, dgamma2(x, mu = 5, sigma2 = 5), type = 'l', ylim = c(0, 0.8),
     xlab = NA, ylab = 'Density', main = 'high \U03BC, high \U03C3\U00B2')
layout(1)
```

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982]. The hypothesis that spatial needs decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022]. Conceptually, the relationship between spatial needs and $\e(R)$ must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see figure \@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].

When $\e(R)$ is high, we expect animals' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. However, as $\e(R)$ decreases, we expect animals' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force animals to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an animal's lifespan. For instance, larger home ranges requires greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

@@

Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat. However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and is lowest in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.

This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions where variability does not change noticeably spatiotemporally, or when data availability is too low to produce appreciable measures of changes in variance. Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case. Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response.

# Effects of resource stochasticity, $\var(R)$

>>>>

---
title: \large How resource abundance and stochasticity affect organisms' spatial needs
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Simon Wood
    institute: uofe
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: E. Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - uofe: School of Mathematics, James Clerk Maxwell Building, University of Edinburgh, Edinburgh, Unted Kingdom.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    # - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

<!-- define custom math symbols -->
\newcommand*\e{\text{E}} <!-- expectation; non-italic E -->
\newcommand*\var{\text{Var}} <!-- variance; non-italic Var -->

@@

#' place figures before the paragraph they are referred to so `wrapfigure`
#' (generally) places them beside the paragraph
```

@@

\noindent The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the *mean* amount of resources, while responses to the variance around the mean are still largely unknown. This is not a trivial oversight. Organisms adjust to variable environmental conditions, so failing to account for the effects of resource unpredictability can result in a limited understanding of organisms' spatial needs, challenging ecological theory and applied conservation alike. In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance. We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on organisms' space use. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in space use. Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals. Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.

@@

The relationship between organisms' movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range. Three decades after, @southwood_habitat_1977 suggested that change in resource abundance is a strong determinant of how organisms decide where to live and when to reproduce. Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an organism's home-range size is

@@

\noindent where $H$ is the organism's **home-range size**, $C$ is the organism's resource consumption, and $R$ is the **resources** the organism can access. Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but a species' spatial requirements likely depend on numerous other factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020]. While many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, few studies build upon others' previous work. Instead, results are reported as independent, novel findings. Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on organisms' spatial needs. In addition, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.

In this paper, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. For instance, a grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources. In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement and life history strategies being selected for. Although it is generally expected that resource unpredictability will decrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

Here, we illustrate how organisms' spatial needs depend on both the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation. Next, we provide a review of the effects of resource *abundance* on organisms' spatial needs while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource *stochasticity* on organisms' spatial needs while suggesting a second unifying hypothesis. Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022]. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

# Simulating resources as a random variable

\noindent In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa). As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically, since resources tend to be less variable when they are less abundant. We assume readers are familiar with the concepts of: random variables, probability distributions, the expected value of a random variable, and the variance of a random variable. We define and explain each of the four concepts in Appendix 1 for readers who need additional information.

## Effects of resource abundance, $\e(R)$

```{r mu-hyp, fig.cap="Hypothesized spatial needs of an organism as a function of resource abundance ($\\e(R)$). We expect low values of $\\e(R)$ to result in high spatial needs as organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic. As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the organism to survive. Note that the relationship between $\\e(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
```

\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982]. The hypothesis that spatial needs decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022]. Conceptually, the relationship between spatial needs and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see figure \@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].

When $\e(R)$ is high, we expect organisms' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. However, as $\e(R)$ decreases, we expect organisms' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force organisms to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on organism's spatial needs and when an organism may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an organism's lifespan. For instance, larger home ranges requires greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

@@

Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat. However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.

This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions where variability does not change noticeably spatiotemporally, or when data availability is too low to produce appreciable measures of changes in variance. Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case. Most often, $\e(R)$ will vary over time, and organisms' behaviors will change in response.

## Effects of resource stochasticity, $\var(R)$

\noindent Although the effect of $R$ on organisms' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977;  @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021]. However, $\var(R)$ can fluctuate temporally (and spatially) due to many important drivers, including repetitive and predictable patterns as well as more stochastic ones. For example, since berries are scarce outside the growing season, the variance in berries will also be low. But during the growing season, both the average number of berries and the variance in berries are higher, since a bear may explore a location before fruiting time or after competitors ate all the berries. In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors [@alexander_novel_2015], the emergence of new diseases [@hollings_trophic_2014], droughts [@foley_severe_2008; @haig_assessment_2013], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], changes in climate and phenology [@berger_climate_2018; @grant_evolution_2017; @jonzen_rapid_2006; @inouye_climate_2000; @severson_spring_2021; @woolway_global_2020], and other extreme events [@logares_black_2012; @anderson_black-swan_2017].

```{r s2-hyp, fig.cap="Hypothesized spatial needs of an organism as a function of resource stochasticity ($\\var(R)$). We expect low values of $\\var(R)$ to result in low spatial needs as organisms are able to depend on relatively predictable resources. As $\\var(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident organisms) or a switch to dispersal, nomadism, or migration. Note that the relationship between $\\var(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
```
